---
title: "RNA-seq Differential Expression Analysis of *Saccharomyces cerevisiae*"
author: "Songyuan He"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

## Project Overview

This notebook demonstrates a complete RNA-seq differential expression analysis
pipeline applied to a *Saccharomyces cerevisiae* (budding yeast) dataset, as
part of the **Genomics & Bioinformatics** coursework. The analysis covers the
full workflow from raw count matrix to functional interpretation.

**Key packages:** `DESeq2` (differential expression), `clusterProfiler`
(GO enrichment), `ggplot2` (visualisation).

---

## 1. Setup and Dependencies

```{r setup, message=FALSE, warning=FALSE}
# ── CRAN packages ──
library(ggplot2)
library(dplyr)
library(tidyr)
library(pheatmap)

# ── Bioconductor packages ──
library(DESeq2)
library(clusterProfiler)
library(org.Sc.sgd.db)   # S. cerevisiae genome-wide annotation
library(RColorBrewer)

# ── Session info for reproducibility ──
Sys.setlocale("LC_ALL", "en_US.UTF-8")
```

---

## 2. Data Loading

Count matrix and sample metadata are stored under `data/`. The experiment
compares **two control** versus **two treatment** samples (SRA accessions
SRR10720198–SRR10720205).

```{r load-data}
# Read count matrix (genes x samples)
count_data <- as.matrix(read.csv(
  file.path("data", "countData.csv"),
  sep = ",",
  row.names = "gene_id"
))

# Read sample metadata
col_data <- read.csv(
  file.path("data", "colData.csv"),
  sep = ",",
  row.names = 1
)

# ── Sanity checks ──
stopifnot(all(rownames(col_data) %in% colnames(count_data)))
count_data <- count_data[, rownames(col_data)]
stopifnot(all(rownames(col_data) == colnames(count_data)))

cat(sprintf(
  "Loaded %d genes × %d samples (%s)\n",
  nrow(count_data), ncol(count_data),
  paste(colnames(count_data), collapse = ", ")
))
print(col_data)
```

---

## 3. Quality Control

### 3.1 Count Distribution

```{r qc-count-dist, fig.width=8, fig.height=5}
count_long <- as.data.frame(count_data) %>%
  tibble::rownames_to_column("gene_id") %>%
  tidyr::pivot_longer(-gene_id, names_to = "sample", values_to = "count")

ggplot(count_long, aes(x = count + 1, colour = sample)) +
  geom_density(linewidth = 1) +
  scale_x_log10() +
  labs(
    title = "Raw count distribution per sample",
    x = "log10(count + 1)", y = "Density"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")
```

### 3.2 Genes with Zero Counts

```{r qc-zeros}
zero_fraction <- rowMeans(count_data == 0)
cat(sprintf(
  "Genes with zero counts in all samples: %d / %d (%.1f%%)\n",
  sum(zero_fraction == 1), nrow(count_data),
  100 * sum(zero_fraction == 1) / nrow(count_data)
))
```

---

## 4. DESeq2 Differential Expression

### 4.1 Build DESeq2 Dataset

```{r deseq-setup}
dds <- DESeqDataSetFromMatrix(
  countData = count_data,
  colData   = col_data,
  design    = ~ type
)

# Pre-filter: keep genes with ≥ 10 reads total
dds <- dds[rowSums(counts(dds)) >= 10, ]
cat(sprintf("Genes retained after pre-filtering: %d\n", nrow(dds)))
```

### 4.2 Run DESeq2

```{r deseq-run}
dds <- DESeq(dds)

# Extract results: treat vs control
res <- results(dds, contrast = c("type", "treat", "control"), alpha = 0.05)
res <- res[order(res$pvalue), ]
res <- as.data.frame(res)

# Summary
summary(res)
```

```{r deseq-summary-table}
# Display top 10 DEGs by adjusted p-value
res_sig <- subset(res, padj < 0.05)
cat(sprintf(
  "Differentially expressed genes (padj < 0.05): %d (up: %d, down: %d)\n",
  nrow(res_sig),
  sum(res_sig$log2FoldChange > 0, na.rm = TRUE),
  sum(res_sig$log2FoldChange < 0, na.rm = TRUE)
))
knitr::kable(head(res_sig[order(res_sig$padj), ], 10),
             caption = "Top 10 DEGs by adjusted p-value")
```

---

## 5. Visualisation

### 5.1 Volcano Plot

```{r volcano, fig.width=8, fig.height=6}
res_plot <- res
res_plot$significance <- "Not significant"
res_plot$significance[res_plot$padj < 0.05 & res_plot$log2FoldChange > 1]  <- "Up"
res_plot$significance[res_plot$padj < 0.05 & res_plot$log2FoldChange < -1] <- "Down"

ggplot(res_plot, aes(x = log2FoldChange, y = -log10(pvalue), colour = significance)) +
  geom_point(alpha = 0.5, size = 1.5) +
  scale_colour_manual(values = c("Up" = "#E64B35", "Down" = "#4DBBD5", "Not significant" = "grey70")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey40") +
  labs(
    title = "Volcano plot: treat vs control",
    x = expression(log[2]~fold~change),
    y = expression(-log[10]~italic(p))
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")
```

### 5.2 MA Plot

```{r ma-plot, fig.width=8, fig.height=6}
plotMA(res, alpha = 0.05, main = "MA plot: treat vs control", ylim = c(-5, 5))
```

### 5.3 Heatmap of Top Differentially Expressed Genes

```{r heatmap, fig.width=8, fig.height=8}
top_genes <- head(rownames(res_sig[order(res_sig$padj), ]), 30)

# Variance-stabilising transformation for visualisation
vsd <- vst(dds, blind = FALSE)
mat <- assay(vsd)[top_genes, ]
rownames(mat) <- top_genes

# Z-score normalisation per gene
mat_scaled <- t(scale(t(mat)))

pheatmap(
  mat_scaled,
  cluster_rows    = TRUE,
  cluster_cols    = TRUE,
  annotation_col  = as.data.frame(colData(dds)["type"]),
  main            = "Top 30 DEGs (Z-score-normalised VST counts)",
  color           = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100),
  fontsize_row    = 7,
  border_color    = NA
)
```

### 5.4 PCA Plot

```{r pca, fig.width=7, fig.height=6}
plotPCA(vsd, intgroup = "type") +
  geom_point(size = 4) +
  labs(title = "PCA of variance-stabilised transformed counts") +
  theme_bw(base_size = 13)
```

---

## 6. Functional Enrichment (clusterProfiler)

Gene Ontology enrichment for differentially expressed genes across all three
GO categories: Biological Process (BP), Cellular Component (CC), and Molecular
Function (MF).

```{r go-enrichment, fig.width=10, fig.height=7}
sig_genes <- rownames(res_sig)

go_bp <- enrichGO(
  gene          = sig_genes,
  OrgDb         = org.Sc.sgd.db,
  keyType       = "ORF",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05
)

go_cc <- enrichGO(
  gene          = sig_genes,
  OrgDb         = org.Sc.sgd.db,
  keyType       = "ORF",
  ont           = "CC",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05
)

go_mf <- enrichGO(
  gene          = sig_genes,
  OrgDb         = org.Sc.sgd.db,
  keyType       = "ORF",
  ont           = "MF",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05
)

# ── Dotplot of top 10 terms per category ──
go_list <- list(BP = go_bp, CC = go_cc, MF = go_mf)
go_list <- go_list[!sapply(go_list, is.null)]

if (length(go_list) > 0) {
  dotplot_list <- lapply(names(go_list), function(ont) {
    go_res <- go_list[[ont]]
    if (!is.null(go_res)) {
      p <- dotplot(go_res, showCategory = 10, title = ont)
      p + theme_bw(base_size = 11)
    }
  })
  patchwork::wrap_plots(dotplot_list, ncol = 1)
}
```

```{r go-export}
# Export formatted results
export_go <- function(go_res, category_label) {
  if (is.null(go_res)) return(NULL)
  df <- as.data.frame(go_res)
  df$Category <- category_label
  df <- df %>% arrange(p.adjust) %>% head(10)
  df[, c("Category", "Description", "GeneRatio", "BgRatio", "pvalue", "p.adjust", "Count")]
}

go_all <- rbind(
  export_go(go_bp, "Biological Process"),
  export_go(go_cc, "Cellular Component"),
  export_go(go_mf, "Molecular Function")
)
write.csv(go_all, file.path("data", "GO_enrichment_results.csv"), row.names = FALSE)
cat("GO enrichment results saved to data/GO_enrichment_results.csv\n")
```

---

## 7. Session Information

```{r session-info}
sessionInfo()
```
